Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2019-11-16 manuscript/climate_sensitivity_draft.Rmd
loaded Bill's feedback
plust created folder, added Bill & Norm as definite authors
Consolidate author feedback - up to end of 2nd paragraph of intro
#65

Also consolidated all author feedback into one document (minus Zailaa)

**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

**Authors:** 
Ian R. McGregor^1,2^
Ryan Helcoski^1^
Norbert Kunert^1,3^
Alan J. Tepley^1,4^
Erika B. Gonzalez-Akre^1^
Valentine Herrmann^1^
Joseph Zailaa^1,5^
Atticus Stovall^1,6^
Norman A. Bourg^1^?
William J. McShea^1^?
Neil Pederson^7^
Lawren Sack^8^
Kristina J. Anderson-Teixeira^1,3^*

@@

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
2.
3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
4.
5. Biological Sciences Department; California State University; Los Angeles, CA, USA
6.
7.
8.

@@

text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	currently 200 (limit 200) | 	No. of Tables | 	5  
Introduction|	currently ~1339 |	No of Supporting Information files | 	6
Materials and Methods |	currently ~1604	||	 
Results |	currently ~1558	 ||
Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | currently ~73	||

@@

- Individual-level drought resistance decreased with tree height, whereas crown position explained little additional variation and root water access did not confer additional resistance. Resistance was lower among species whose leaves had higher percent loss of area upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across given droughts.

@@

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these.*

\newpage
### Introduction
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014], sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Global change-type droughts have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests around the world [@allen_global_2010; **REFS**]. During drought, larger trees generally suffer greater reductions in growth and increases in mortality [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic meaning. This study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees. First, taller trees face greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], which becomes a greater liability during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because they are more often in the canopy, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; **REFS-KAT**). Alternatively, the generally supressed status of subcanopy trees may be insufficient to override the benefits of their buffered environment during drought. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. A final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience. Alternatively, larger size classes may be dominated by species that are better adapted to inherently greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms underlying the tendency for larger trees to suffer more during drought will require sorting out the interactive effects of height, canopy position, root water acess, and species' traits.

>>>>

**Title:** Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf deciduous forest

**Authors:** 
Ian R. McGregor^1,2^,
Ryan Helcoski^1^,
Norbert Kunert^1,3^,
Alan J. Tepley^1,4^,
Erika B. Gonzalez-Akre^1^,
Valentine Herrmann^1^,
Joseph Zailaa^1,5^,
Atticus Stovall^1,6^,
Norman A. Bourg^1^,
William J. McShea^1^,
Neil Pederson^7^,
Lawren Sack^8^,
Kristina J. Anderson-Teixeira^1,3^*

@@

1. Conservation Ecology Center; Smithsonian Conservation Biology Institute; National Zoological Park, Front Royal, VA, USA
2.
3. Center for Tropical Forest Science-Forest Global Earth Observatory; Smithsonian Tropical Research Institute; Panama, Republic of Panama
4. Canadian Forest Service, Northern Forestry Centre, Edmonton, Alberta, Canada
5. Biological Sciences Department; California State University; Los Angeles, CA, USA
6.
7.
8.

@@

Text | word count | other | n
---------------------------------------------|---------------------|-----------------------------|--------------------
Total word count (excluding summary, references and legends) |  currently ~6442 (strict limit 6,500)	 |	No. of figures | 	2 (both colour)
Summary|  	currently 200 (limit 200) | 	No. of Tables | 	5  
Introduction|	currently ~1339 |	No of Supporting Information files | 	6
Materials and Methods |	currently ~1604	||	 
Results |	currently ~1558	 ||
Discussion |	currently ~1866 (limit 30% of total (not strict), or 1950 if manuscript reaches word limit)	||
Acknowledgements | currently ~73	||

@@

- Individual-level drought resistance decreased with tree height, whereas crown position and root water access explained little variation. Resistance was lower among species whose leaves had a higher percent loss of area upon dehydration and lost turgor at less negative water potential. However, there was substantial variation in the best predictor variables across the three drought periods. **I think Alan's comments here warrant discussion. I believe the answer to his second comment is yes. In addition, Erika suggests spelling out the best predictor traits despite the point below. Thoughts?**

@@

*Key words*:  canopy position; drought; Forest Global Earth Observatory (ForestGEO); leaf hydraulic traits; temperate broadleaf deciduous forest; tree growth; tree height; tree-ring [5-8] *Five to eight key words (in alphabetical order) . Words that are in the title can, and should, be among these.* **Erika is of the opinion that broadleaf should not be included here, instead using "deciduous". She also added deciduous to the title. I'm unsure if this is redundant**

\newpage
### Introduction
**Nobby says this intro is based too much on Bennet 2015. I'm not sure if he means we should remove some of the citations or rephrase some sentences.**
Forests globally play a critical role in climate regulation [@bonan_forests_2008], yet there remains enormous uncertainty as to how the terrestrial carbon (C) sink, which is dominated by forests, will respond to climate change [@friedlingstein_climatecarbon_2006]. An important aspect of this uncertainty lies with physiological responses of trees to drought [@kennedy_implementing_2019]. In many forested regions around the world, the risk of severe drought is increasing [@trenberth_global_2014] **Neil adds in a number of refs here, unsure if we should just include all of them**, sometimes despite increasing precipitation [@intergovernmental_panel_on_climate_change_climate_2015]. Droughts resulting from global change have been affecting forests worldwide [@allen_global_2010], and are expected to continue severely impacting forests [@allen_global_2010; **REFS**] **Alan suggests not citing Allen2010 here and instead gives a list of other sources. Erika also suggests 2 papers**. During drought, larger trees generally suffer greater reductions in growth and increased mortality rates [e.g., @bennett_larger_2015; @stovall_tree_2019], resulting in disproportionate impacts on forest C storage [@meakem_role_2018]. As a result, forest drought responses stand to strongly impact forest feedbacks to climate change [@mcdowell_multi-scale_2016; @charney_observed_2016]. Yet, accurate characterization of drought responses remains a modeling challenge [@kennedy_implementing_2019]-- in part because some of the underlying mechanisms remain unclear (**REF**). Understanding forest responses to drought requires increased functional understanding of how tree size, microhabitat, and species' traits jointly confer individual-level vulnerability or resistance, and the extent to which their influence is consistent across droughts.

One fundamental question regarding forest responses to drought is what drives the observed tendency for large trees to suffer more during drought. @bennett_larger_2015 was the first to show larger growth reductions for larger trees on a global scale, and numerous subsequent studies have reinforced this finding (e.g., @stovall_tree_2019; @hacket-pain_consistent_2016]. However, @bennett_larger_2015 quantified tree size based on DBH, which has no direct mechanistic response. That study proposed two major mechanisms--besides the tendency for bark beetles to preferentially attack larger trees [@pfeifer_observations_2011]--for the observed greater drought growth reductions of large trees **Alan gives another mechanism here from another paper. Bill says this sentence is confusing but I believe it is fine**. First, they found taller trees have a greater biophysical challenge of lifting water greater distances against the effects of gravity and friction [@mcdowell_relationships_2011; @mcdowell_darcys_2015; @ryan_hydraulic_2006; @couvreur_water_2018], especially during drought [@zhang_size-dependent_2009]. Second, larger trees may have lower drought resistance because of their higher crown positions, where they are exposed to higher solar radiation, greater wind speeds, lower humidity, and lower CO_2_ concentrations [@koike_leaf_2001; @kunert_revised_2017 **REFS-KAT**). Subcanopy trees subsequently tend to fare better specifically due to the benefits of a buffered environment [@pretzsch_drought_2018]. Potentially counteracting the biophysical challenges faced by large trees, their larger root systems confer an advantage in terms of allowing greater access to water; however, it appears that this effect is usually insufficient to offset the costs of height and/or crown exposure [@bennett_larger_2015]. **For this previous sentence, Alan questions whether this assumption can be made, and Valentine says it's odd to come back to Bennett as a ref here.** In addition to what was described by @bennett_larger_2015, a final mechanism that could mediate tree size-related responses to drought is how hydraulic traits are distributed with respect to size [@meakem_role_2018; @liu_hydraulic_2019]. The pattern observed by @bennett_larger_2015 could be caused if the larger size classes were dominated by species less adapted to handle drought, be it through avoidance, resistance, or resilience **Valentine is confused if you mean to say "lack of" these things. Alan points out we should define these terms here since is the first time they're used. Unsure what your thoughts are on that**. Alternatively, larger size classes may be dominated by species that are inherently better adapted to greater biophysical challenges--as is the case in tropical moist forests of Panama, where larger size classes contain greater proportions of deciduous species [@condit_quantifying_2000; @meakem_role_2018]. Understanding the mechanisms driving more severe suffering of larger trees during drought will require sorting out the interactive effects of height, canopy position, root water access, and species' traits.
